On the surprising effectiveness of a simple matrix exponential derivative approximation, with application to global SARS-CoV-2

### 2023-10-25 manuscript/pnas_revision.tex
revision done question mark

@@

\begin{figure}[!ht]
	\centering
\includegraphics[width=\linewidth]{lowDimVis.pdf}
\vspace{-2em}
\caption{Posterior density plots for surrogate-trajectory Hamiltonian Monte Carlo (HMC) (\cref{sec:Sur:trad:HMC}) using the naive approximate derivative, the corrected approximation \eqref{eq:better:help:on:sale} and the exact matrix exponential derivative for elements of a $5\times 5$ generator matrix. The near perfect overlap reflects the fact that each algorithm's transition kernel leaves the posterior distribution invariant \cite{glatt2020accept}.  To generate data, we randomly draw standard normal generator entries once and simulate 20 independent initial/final position pairs from a CTMC with time interval $t=1$.  We show the true value in red and negate diagonal elements to simplify presentation.  }\label{fig:lowDim}
\end{figure}

@@

>>>>

\begin{figure}[!t]
	\centering
	\includegraphics[width=\linewidth]{lowDimVis.pdf}
	\vspace{-2em}
	\caption{Posterior density plots for surrogate-trajectory Hamiltonian Monte Carlo (HMC) (\cref{sec:Sur:trad:HMC}) using the naive approximate derivative, the corrected approximation \eqref{eq:better:help:on:sale} and the exact matrix exponential derivative for elements of a $5\times 5$ generator matrix. The near perfect overlap reflects the fact that each algorithm's transition kernel leaves the posterior distribution invariant \cite{glatt2020accept}.  To generate data, we randomly draw standard normal generator entries once and simulate 20 independent initial/final position pairs from a CTMC with time interval $t=1$.  We show the true value in red and negate diagonal elements to simplify presentation.  }\label{fig:lowDim}
\end{figure}

@@

@@

Here, we fill in remaining simulation details not included in figure captions. In the simulations contributing to \cref{fig:app}, we randomly generate new, independent direction matrices $\JJ$ at each time step within each of the 20 independent runs. The results are not sensitive to $\JJ$ in general.  We also note that under \cref{def:sub:exp:RV} the Cauchy distribution is \emph{not} sub-exponential and therefore represents a deviation from core assumptions of \cref{sec:rig:results}.  Using varying distributions on the elements of generator matrices $\QQ$, the simulations contributing to \cref{fig:lowDim} and \cref{fig:truthMean} both randomly generate $N$ independent initial states $\y^0_1,\dots,\y^0_N$ according to uniform distributions over their respective CTMC state spaces. For each of these initial states $\y_0$, we then simulate from the CTMC for time interval $t=1$ to obtain samples $\y^1_1,\dots,\y^1_N$.  The likelihood then takes the form $\prod_n (\y^{0}_n)^T e^{\QQ} \y^{1}_n$.  For the simulation contributing to \cref{fig:lowDim}, we specify standard normal priors and sample from the posterior by generating 100,000 iterations from each algorithm.  For the simulation contributing to \cref{fig:truthMean}, we generate 500,000 MCMC samples using surrogate-trajectory HMC with the naive matrix exponential derivative and calculate the posterior mean of each generator matrix element using the final 100,000 samples.
